How to Avoid OOTA Without Really Trying

### 2024-09-13 paper/ootavacuous.tex
treewide: Use "threads" and "Tn" to align with C++ nomenclature
No more "processes" named "Pn".  Also fix a grammar nit while in the
area.

Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{listing}[bp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}

@@

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote private per-process variables
	(``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, process P0's line~1 does a load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, P1's line~2 stores 42 to \co{x}, justifying the value loaded
by P0's line~1.

@@

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
P0 and P1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	A similar restriction applies to compiler-based speculation.}
the stores on line~2 cannot take place until the P0's and P1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

\begin{listing}[bp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = 42} \\
\end{tabular}

@@

Simple reordering can cause non-OOTA, but OOTA-like, behavior, as
exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Either the compiler or the CPU might reorder P1's lines~1 and~2, so that
all of \co{x}, \co{y}, \co{r1}, and \co{r2} might end up with the value 42,
as indicated by the ``sometimes satisfied'' in the figure.

@@

This is a fine definition and is consistent with the words in the C++
standard, but it has a problem with intrathread rfi links
as exemplified by the following code: \\
% \begin{quote}
% Constructed by hand.
{
\small
\begin{tabular}{r|l}
% & \multicolumn{1}{c}{P0} \\
% \hline
1. &
\texttt{~~int r2 = x} \\
2. &
\texttt{~~y = r2} \\
3. &
\texttt{~~int r3 = y} \\
4. &
\texttt{~~z = r3} \\
\end{tabular}
} \\
% \end{quote}
This is an elaboration of P1 from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~3 and~4.
The problem is that a C++ implementation may
note that line~3 could well execute immediately after line~2, giving
other threads no chance to modify \co{y} in between.
Such an implementation might therefore behave as if line~3 had been
removed from the source code and line~4's \co{r3} had been replaced by
\co{r2}.
This means that line~4 stores the same value to \co{z} that was stored
to \co{y} by line~2.
And given that the load from \co{y} no longer exists, it cannot possibly
act as a temporal constraint.

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which P0 loads the value of \co{x} and stores it in \co{y} while P1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code: \\
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~z = r2} \\
\end{tabular}
} \\
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

A CPU may start executing an instruction speculatively, but at some
point it must either abandon that instruction or \emph{commit} it with
a set of specific outputs based on well-defined inputs.
A CPU is clearly not permitted to commit an instruction its inputs have
been committed, unless the remaining uncommitted inputs cannot affect
that instruction's outputs.
For instance, a conditional-move instruction needn't wait for the source
input to commit if it knows that the condition is definitely false.

@@

For example, consider the Simple OOTA program shown in
Listing~\ref{lst:Simple OOTA}.
A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}''.
This would be a valid transformation, and the resulting behavior would
not count as OOTA according to our definition because the
dependencies in P0 and P1 would not be semantic.

>>>>

\begin{listing}[bp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}

@@

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote private per-process variables
	(``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, thread T0's line~1 does a load from \co{x} into \co{r1},
reading the value of T1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, T0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, T1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, T1's line~2 stores 42 to \co{x}, justifying the value loaded
by T0's line~1.

@@

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
T0 and T1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	A similar restriction applies to compiler-based speculation.}
the stores on line~2 cannot take place until the T0's and T1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

\begin{listing}[bp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = 42} \\
\end{tabular}

@@

Simple reordering can cause non-OOTA, but OOTA-like, behavior, as
exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Either the compiler or the CPU might reorder T1's lines~1 and~2, so that
all of \co{x}, \co{y}, \co{r1}, and \co{r2} might end up with the value 42,
as indicated by the ``sometimes satisfied'' in the figure.

@@

This is a fine definition and is consistent with the words in the C++
standard, but it has a problem with intrathread rfi links
as exemplified by the following code: \\
% \begin{quote}
% Constructed by hand.
{
\small
\begin{tabular}{r|l}
% & \multicolumn{1}{c}{T0} \\
% \hline
1. &
\texttt{~~int r2 = x} \\
2. &
\texttt{~~y = r2} \\
3. &
\texttt{~~int r3 = y} \\
4. &
\texttt{~~z = r3} \\
\end{tabular}
} \\
% \end{quote}
This is an elaboration of T1 from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~3 and~4.
The problem is that a C++ implementation may
note that line~3 could well execute immediately after line~2, giving
other threads no chance to modify \co{y} in between.
Such an implementation might therefore behave as if line~3 had been
removed from the source code and line~4's \co{r3} had been replaced by
\co{r2}.
This means that line~4 stores the same value to \co{z} that was stored
to \co{y} by line~2.
And given that the load from \co{y} no longer exists, it cannot possibly
act as a temporal constraint.

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code: \\
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~z = r2} \\
\end{tabular}
} \\
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

A CPU may start executing an instruction speculatively, but at some
point it must either abandon that instruction or \emph{commit} it with
a set of specific outputs based on well-defined inputs.
A CPU is clearly not permitted to commit an instruction until its inputs
have been committed, unless the remaining uncommitted inputs cannot
affect that instruction's outputs.
For instance, a conditional-move instruction needn't wait for the source
input to commit if it knows that the condition is definitely false.

@@

For example, consider the Simple OOTA program shown in
Listing~\ref{lst:Simple OOTA}.
A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}''.
This would be a valid transformation, and the resulting behavior would
not count as OOTA according to our definition because the
dependencies in T0 and T1 would not be semantic.
